---
title: "Nested cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A model tuned by cross-validation comes with a score for the winning
candidate, and that candidate was kept because it scored best, so the score
overstates what the model will do on new data; `vignette("estimate")` makes
the case for nesting and says what the nested number means.

Nested cross-validation puts the whole tune-and-fit procedure inside an outer
resampling loop. Each outer fold tunes on its own analysis rows, fits the
winner there, and scores it once on assessment rows the tuning never saw. The
mean of those outer scores is the estimate.

The number `collect_metrics()` reports describes the whole tune-and-fit
procedure: resample, tune, select, fit. That is the number to report. The
model fitted at the end for deployment is a separate object, and it has no
performance number of its own, because everything computable from its
training data was used up in selecting and fitting it.

This page walks from a design to a write-up. `vignette("results")` reads
everything the results object holds, `vignette("tuners")` swaps in the other
inner searches and scores a workflow with nothing to tune, and the
[parallel article](https://nestedtune.tidymodels.org/articles/parallel.html)
runs the same call on a pool of workers.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This guide attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this guide has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design

`nested_resamples()` builds the two-level structure: an outer resampling,
with an inner resampling attached to each outer fold.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

folds
```

Each row is one outer fold. `splits` holds the outer split, and
`inner_resamples` holds an ordinary `rset` built from that fold's analysis
rows alone, which is all the tuning for that fold gets to see.

```{r inner}
folds$inner_resamples[[1]]
```

`mtcars` has `r nrow(mtcars)` rows, which keeps this page fast to build.
Nesting removes the most bias on wide data searched hard, and
`vignette("estimate")` says when it is worth the cost; this example is here
because it builds in seconds.

## The model and the grid

Anything tune can tune, this can tune. Here it is a random forest with two
parameters marked for tuning, and an explicit grid of candidates.

```{r workflow}
rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
grid
```

That is `r nrow(grid)` candidates, each resampled inside every outer fold.
The run below fits `r nrow(folds)` outer folds times
`r nrow(folds$inner_resamples[[1]])` inner resamples times `r nrow(grid)`
candidates, `r nrow(folds) * nrow(folds$inner_resamples[[1]]) * nrow(grid)`
models for tuning, plus one per outer fold for scoring. Nested
cross-validation is expensive, and that product is where the cost lives.

## Running the loop

`nested_tune_grid()` drives the outer loop. For each outer fold it calls
`tune::tune_grid()` on that fold's inner resamples, selects a candidate by
the `select` rule (the best by the first metric unless `selection_rule()`
says otherwise), finalizes the workflow, and fits and scores it on the outer
split. Every statistical step is tune's. This package adds the loop, the
seeding, and a result that keeps what each fold chose.

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

Printing shows the object: one row per outer fold, and the record each fold
left behind. `summary()` says what the run means.

```{r summarize}
summary(res)
```

## The number to report

```{r estimate}
est <- collect_metrics(res)
est

rmse_row <- filter(est, .metric == "rmse")
```

The RMSE of `r round(rmse_row$mean, 2)` is the mean of `r rmse_row$n`
scores, each measured on rows the procedure never touched. `std_err` is the
standard error of that mean, not the spread of the folds and not a
confidence interval. Two of these numbers from two workflows cannot be
subtracted to compare them, and `vignette("estimate")` says why.

## Each fold's selection

The summary above listed what each fold selected, and `.selected` is where
those choices live: a list column of one-row tibbles, one per outer fold.
`collect_selections()` stacks them, with the fold each came from beside it.

```{r selected}
selected <- collect_selections(res)

selected
```

```{r instability}
n_mtry <- n_distinct(selected$mtry)
n_min_n <- n_distinct(selected$min_n)
```

Across `r nrow(res)` outer folds, `mtry` took `r n_mtry` distinct selected
values and `min_n` took `r n_min_n`. Most tools throw this away. nestedtune
keeps it, because it is information about the procedure rather than noise in
it.

`autoplot()` draws those same selections, one panel per tuned parameter and
one point per outer fold. A flat row means the folds agreed. Scatter means
they did not.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

A parameter the folds agree on is one the data picks clearly. A parameter
they split over is one whose value is largely arbitrary at this sample size,
which is expected wherever the candidates perform about equally well;
`vignette("estimate")` gives the mechanism.

The per-fold scores are worth a look for the same reason:

```{r per-fold}
per_fold <- collect_metrics(res, summarize = FALSE)
per_fold

fold_rmse <- per_fold |>
  filter(.metric == "rmse") |>
  pull(.estimate)

c(sd = sd(fold_rmse), std_err = sd(fold_rmse) / sqrt(length(fold_rmse)))
```

The first number is how much the folds differ from each other, and the
second is the `std_err` that `collect_metrics()` reports, the precision of
their mean. Quoting the second as though it described the folds understates
their disagreement. The other `autoplot()` view draws that spread, with a
dashed line at the nested estimate.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, with one point per outer fold's score and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

A tuned procedure is usually compared with something simpler, such as the
same model with its parameters fixed, and `vignette("tuners")` scores such a
workflow on these same folds with `nested_fit_resamples()`.

## The model to deploy

Nothing above produced a model to predict with, and that is deliberate. The
model is built by running the same procedure once more with the whole
dataset in hand. The procedure is read from `res` (the inner resampling
specification, the grid, the metrics, the selection rule), so the model and
the estimate come from one search.

```{r final}
set.seed(3)

final <- nested_final_fit(wf, res)

final
```

The outer folds play no part here. Their selections are not pooled or voted
on; they belong to the estimate.

The object predicts directly. `predict()` and `augment()` on it are the
trained workflow's own methods, and `extract_workflow()` returns the workflow
itself.

```{r predict}
predict(final, new_data = mtcars[1:3, ])
```

`extract_tune_results()` returns the tuning run this model's parameters came
from, an ordinary tune result, and `show_best()` on it gives the number most
tempting to report:

```{r comparison}
best_selection <- extract_tune_results(final) |>
  show_best(metric = "rmse", n = 1)

best_selection

tibble(
  quantity = c("nested estimate (report this)", "best selection-time score"),
  rmse = c(rmse_row$mean, best_selection$mean)
)
```

The selection-time score was computed on the resamples that chose the
winner, so it is not an estimate of performance on anything, whichever side
of the nested estimate it lands on; `vignette("estimate")` says why. It
stays on the object as the record of what selection saw, and tune's readers
hand it over without a warning. That is why the results object and the final
fit both refuse tune's ranking generics rather than answering them. On the
loop's results they would rank outer folds, and on the final fit there is
one model and nothing to rank.

```{r refusals, error = TRUE}
show_best(res, metric = "rmse")
```

```{r refusals-final, error = TRUE}
select_best(final, metric = "rmse")
```

## Reproducibility

Seed the session before the call, as elsewhere in tidymodels; neither
function takes a seed of its own. Each draws a tuning seed and an outer-fit
seed for every fold from the session state at entry, fixed by the fold's
position in the design rather than by the order the folds happen to run in,
and stores them on the result. So the same seed gives the same answer whether
the folds run serially or on a pool of workers, and any one fold can be
reproduced by hand. The final fit carries the two seeds it ran under as
`tuning_seed` and `fit_seed`, and the caller's own random state is put back
as it was found.

```{r seeds}
res$.tuning_seed
final$tuning_seed
```

```{r rng-restored}
before <- .Random.seed

invisible(nested_final_fit(wf, res))

identical(before, .Random.seed)
```

`?nested_tune_grid` gives the recipe for reproducing one fold by hand, and
`?nested_final_fit` the one for redoing the final fit from its two seeds.

## The write-up

Everything a write-up needs is on the two objects. A minimal, honest report:

```{r writeup, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "> Hyperparameters (`mtry`, `min_n`) were tuned over a %d-point grid by ",
    "%d-fold cross-validation, nested inside a %d-fold outer cross-validation ",
    "of the entire tune-and-fit procedure (n = %d). The outer folds give an ",
    "estimated RMSE of %.2f (SE %.2f) for the procedure. Across those %d ",
    "folds, selection took %d distinct values of `mtry` and %d of `min_n`. ",
    "The deployed model was produced by applying the same procedure to the ",
    "full dataset, which selected mtry = %d and min_n = %d."
  ),
  nrow(grid),
  nrow(folds$inner_resamples[[1]]),
  nrow(folds),
  nrow(mtcars),
  rmse_row$mean,
  rmse_row$std_err,
  nrow(res),
  n_mtry,
  n_min_n,
  final$selected$mtry,
  final$selected$min_n
))
```

Three things make it honest. The estimate is attributed to the procedure and
not to the model, the instability is reported rather than hidden, and the
deployed model is described as what it is: the same procedure applied to all
the data, carrying no performance claim of its own.
