---
title: "Nested cross-validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested cross-validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When you tune a model with cross-validation, you try several settings and
keep the one with the best score. That score is optimistic. You picked the
winner because it scored well, so some of its score is luck, and the same
number will not hold up on new data. `vignette("estimate")` explains why in
more detail.

Nested cross-validation fixes this by scoring the whole tuning procedure
rather than the winner alone. It splits the data into outer folds. Within
each outer fold it tunes using only the analysis rows, fits the winning
setting on those rows, and scores that fit on the assessment rows the
tuning never saw. The average of those outer scores is an honest estimate of
how well tune-then-fit works on new data.

That average is what `collect_metrics()` reports, and it is the number to put
in a write-up. The model you fit at the end on all the data is a separate
object. It has no score of its own, because every row it was trained on was
already used to choose and fit it.

This page walks from a design to a write-up. The other pages go deeper.
`vignette("results")` reads everything the results object holds.
`vignette("tuners")` swaps in the other inner searches and scores a workflow
with nothing to tune. The
[parallel article](https://nestedtune.tidymodels.org/articles/parallel.html)
runs the same call on a pool of workers.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This guide attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this guide has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design

`nested_resamples()` builds the two-level structure: an outer resampling,
with an inner resampling attached to each outer fold.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

folds
```

Each row is one outer fold. `splits` holds the outer split, and
`inner_resamples` holds an ordinary `rset` built from that fold's analysis
rows alone, which is all the tuning for that fold gets to see.

```{r inner}
folds$inner_resamples[[1]]
```

`mtcars` has `r nrow(mtcars)` rows, which keeps this page fast to build.
This example is here because it builds in seconds, not because it needs
nesting. Nesting removes the most bias when the data has many more columns
than rows and the grid is large. `vignette("estimate")` says when it is
worth the cost.

## The model and the grid

Anything tune can tune, this can tune. Here it is a random forest with two
parameters marked for tuning, and an explicit grid of candidates, the
settings to try.

```{r workflow}
rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
grid
```

That is `r nrow(grid)` candidates, each resampled inside every outer fold.
The run below has `r nrow(folds)` outer folds. Each outer fold tunes on
`r nrow(folds$inner_resamples[[1]])` inner resamples, and each inner
resample scores all `r nrow(grid)` candidates. That multiplies out to
`r nrow(folds) * nrow(folds$inner_resamples[[1]]) * nrow(grid)` models for
tuning, plus one per outer fold for scoring. Nested cross-validation is
expensive, and that product is where the cost lives.

## Running the loop

`nested_tune_grid()` drives the outer loop. For each outer fold it calls
`tune::tune_grid()` on that fold's inner resamples, the call you would make
by hand, with parallelism off and the fold's seed set. It then picks the
best candidate by the first metric. If you
would rather pick by another of tune's selectors, the `select` argument
takes a `selection_rule()` naming it. It finalizes the
workflow with that candidate, fits it on the fold's analysis rows, and
scores it on the assessment rows. Every statistical step is tune's. This
package adds the loop, the seeding, and a result that keeps what each fold
chose.

```{r run}
set.seed(2)

res <- nested_tune_grid(wf, folds, grid = grid)

res
```

Printing shows the object: one row per outer fold, and the record each fold
left behind. `summary()` says what the run means.

```{r summarize}
summary(res)
```

## The number to report

```{r estimate}
est <- collect_metrics(res)
est

rmse_row <- filter(est, .metric == "rmse")
```

The RMSE of `r round(rmse_row$mean, 2)` is the mean of `r rmse_row$n`
scores, each measured on rows the procedure never touched. `std_err` is the
standard error of that mean, not the spread of the folds and not a
confidence interval. Two of these numbers from two workflows cannot be
subtracted to compare them, and `vignette("estimate")` says why.

## Each fold's selection

The summary above listed what each fold selected, and `.selected` is where
those choices live: a list column of one-row tibbles, one per outer fold.
`collect_selections()` stacks them, with the fold each came from beside it.

```{r selected}
selected <- collect_selections(res)

selected
```

```{r instability}
n_mtry <- n_distinct(selected$mtry)
n_min_n <- n_distinct(selected$min_n)
```

Across `r nrow(res)` outer folds, `mtry` took `r n_mtry` distinct selected
values and `min_n` took `r n_min_n`. Most tools throw this away. nestedtune
keeps it, because it is information about the procedure rather than noise in
it.

`autoplot()` draws those same selections, one panel per tuned parameter and
one point per outer fold. A flat row means the folds agreed. Scatter means
they did not.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

A parameter the folds agree on is one the data picks clearly. A parameter
they split over is one whose value is largely arbitrary at this sample size.
That is expected wherever the candidates perform about equally well, and
`vignette("estimate")` gives the mechanism.

The per-fold scores are worth a look for the same reason:

```{r per-fold}
per_fold <- collect_metrics(res, summarize = FALSE)
per_fold

fold_rmse <- per_fold |>
  filter(.metric == "rmse") |>
  pull(.estimate)

c(sd = sd(fold_rmse), std_err = sd(fold_rmse) / sqrt(length(fold_rmse)))
```

The first number is how much the folds differ from each other. The second
is the `std_err` that `collect_metrics()` reports, which is the precision
of their mean. Quoting the second as though it described the folds understates
their disagreement. The other `autoplot()` view draws that spread, with a
dashed line at the nested estimate.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, with one point per outer fold's score and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

A tuned procedure is usually compared with something simpler, such as the
same model with its parameters fixed. `vignette("tuners")` scores such a
workflow on these same folds with `nested_fit_resamples()`.

## The model to deploy

Nothing above produced a model to predict with, and that is deliberate. The
model is built by running the same procedure once more with the whole
dataset in hand. The procedure is read from `res` (the inner resampling
specification, the grid, the metrics, the selection rule), so the model and
the estimate come from one search.

```{r final}
set.seed(3)

final <- nested_final_fit(wf, res)

final
```

The outer folds play no part here. Their selections are not pooled or voted
on; they belong to the estimate.

The object predicts directly. `predict()` and `augment()` on it are the
trained workflow's own methods, and `extract_workflow()` returns the workflow
itself.

```{r predict}
predict(final, new_data = mtcars[1:3, ])
```

`extract_tune_results()` returns the tuning run this model's parameters came
from, an ordinary tune result, and `show_best()` on it gives the number most
tempting to report:

```{r comparison}
best_selection <- extract_tune_results(final) |>
  show_best(metric = "rmse", n = 1)

best_selection

tibble(
  quantity = c("nested estimate (report this)", "best selection-time score"),
  rmse = c(rmse_row$mean, best_selection$mean)
)
```

The selection-time score is not an estimate of performance on anything.
Whether it lands above or below the nested estimate means nothing either.
It stays on the object as a record of what selection saw. `show_best()`
handed it over above because it was given tune's own object, which does not
warn. Ask the same of `res` or of `final` and both refuse, as the two chunks
below show. On the loop's results, tune's ranking functions would rank outer
folds. On the final fit there is one model and nothing to rank.

```{r refusals, error = TRUE}
show_best(res, metric = "rmse")
```

```{r refusals-final, error = TRUE}
select_best(final, metric = "rmse")
```

## Reproducibility

Seed the session before the call, as elsewhere in tidymodels. Neither
`nested_tune_grid()` nor `nested_final_fit()` takes a seed of its own. At
entry, `nested_tune_grid()` draws two seeds per fold from the session
state. One seeds that fold's tuning and the other its outer fit. Both are
stored on the result. A fold's seeds depend on its
position in the design, not on the order the folds happen to run in. So the
same seed gives the same answer whether the folds run serially or on a pool
of workers. `nested_final_fit()` draws one pair the same way and carries it
as `tuning_seed` and `fit_seed`.
Both functions put your random state back as they found it.

```{r seeds}
res$.tuning_seed
final$tuning_seed
```

```{r rng-restored}
before <- .Random.seed

invisible(nested_final_fit(wf, res))

identical(before, .Random.seed)
```

`?nested_tune_grid` gives the recipe for reproducing one fold by hand, and
`?nested_final_fit` the one for redoing the final fit from its two seeds.

## The write-up

Everything a write-up needs is on the two objects. A minimal, honest report:

```{r writeup, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "> Hyperparameters (`mtry`, `min_n`) were tuned over a %d-point grid by ",
    "%d-fold cross-validation, nested inside a %d-fold outer cross-validation ",
    "of the entire tune-and-fit procedure (n = %d). The outer folds give an ",
    "estimated RMSE of %.2f (SE %.2f) for the procedure. Across those %d ",
    "folds, selection took %d distinct values of `mtry` and %d of `min_n`. ",
    "The deployed model was produced by applying the same procedure to the ",
    "full dataset, which selected mtry = %d and min_n = %d."
  ),
  nrow(grid),
  nrow(folds$inner_resamples[[1]]),
  nrow(folds),
  nrow(mtcars),
  rmse_row$mean,
  rmse_row$std_err,
  nrow(res),
  n_mtry,
  n_min_n,
  final$selected$mtry,
  final$selected$min_n
))
```

Three things make it honest. The estimate is attributed to the procedure and
not to the model. The instability is reported rather than hidden. The
deployed model is described as what it is: the same procedure applied to all
the data, carrying no performance claim of its own.
