---
title: "Reading the results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading the results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Every tuning function in this package returns the same kind of object, a
`nested_results`. The getting-started guide, `vignette("nested-cv")`, prints
it, summarizes it, and reads the estimate, the selections and both plots off
it. This page reads the whole object. It goes through the columns one at a
time, then the readers, the functions such as `collect_metrics()` that turn
the columns into tables. It then shows what a run looks like when one outer
fold fails, and which dplyr verbs keep the class and which shed it.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The run

The design, the workflow and the grid are the guide's, unchanged. There are
five outer folds of `mtcars`, five inner folds under each, and a random
forest with two parameters marked for tuning. The grid's six rows are the
candidates, the settings each fold's inner tuning tries. One thing is
added. A
`control_grid()` with `save_pred = TRUE` asks each fold's outer fit to keep
its predictions, so this page can show that column and the function that
stacks it.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

```{r run}
set.seed(2)

res <- nested_tune_grid(
  wf,
  folds,
  grid = grid,
  control = control_grid(save_pred = TRUE)
)

res
```

## The columns

The object is a tibble with one row per outer fold. The print above shows
the columns that fit its width, and its footer names the rest. Each column
holds one piece of what that fold did.

```{r columns}
names(res)
```

`splits` is the fold's outer split, an ordinary rsample split whose analysis
rows the tuning saw and whose assessment rows scored the fold's model.

```{r splits}
res$splits[[1]]
```

`id` is the fold's label, copied from the design. A repeated design carries
more than one label column, and the object records which columns they are.

```{r id}
res$id
```

`.metrics` is the score of the fold's finalized model on its assessment
rows, one row per metric, in the shape `tune::last_fit()` produces. It is
what `collect_metrics()` averages.

```{r metrics}
res$.metrics[[1]]
```

`.selected` is the candidate the fold's inner tuning chose, a one-row tibble
of the tuned parameters. It is what `collect_selections()`, `summary()`,
`agreement()` and the default `autoplot()` read.

```{r selected}
res$.selected[[1]]
```

`.inner_metrics` is the whole table the inner search scored, every candidate
on every inner resample, averaged. It is what the fold's selection was made
from. `collect_inner_metrics()` stacks it across folds, and
`vignette("tuners")` shows how its shape differs by search.

```{r inner-metrics}
res$.inner_metrics[[1]]
```

`.notes` holds what went wrong inside the fold, in tune's shape with a
`location` naming the stage. A fold that ran cleanly has no rows.

```{r notes}
res$.notes[[1]]
```

`.completed` says whether the fold produced a score. The section on a
failed fold below shows a `FALSE`.

```{r completed}
res$.completed
```

`.tuning_seed` and `.outer_fit_seed` are the two seeds each fold ran under,
drawn at entry and fixed by the fold's position in the design, as
`vignette("nested-cv")` explains under Reproducibility.

```{r seeds}
select(res, .tuning_seed, .outer_fit_seed)
```

`.predictions` is there because the control asked for it. It holds each
completed fold's predictions on its assessment rows, as `tune::last_fit()`
returns them, with `.row` naming the row of the data. A run under the
default control has no such column. A run whose control sets `extract` to a
function has an `.extracts` column beside it, holding what that function
returned for each fold's fitted workflow. A fold that failed holds `NULL`
in either.

```{r predictions}
res$.predictions[[1]]
```

The description of the run itself is stored on the object as attributes
rather than columns. `extract_procedure()` returns the procedure, the record
of what ran: which tuner, that tuner's own arguments such as the grid, and
the control as it took effect. `nested_final_fit()` reruns what that record
describes on the whole dataset.

## The readers

Printing describes the object in hand. `summary()` says what the run means:
how much of the design ran, what each fold chose, and the estimate.

```{r summary}
summary(res)
```

`collect_metrics()` is the estimate as a table, the mean and standard error
of the outer scores. With `summarize = FALSE` it is the outer scores
themselves, one row per fold and metric, with the fold label beside them.

```{r collect-metrics}
collect_metrics(res)

collect_metrics(res, summarize = FALSE)
```

`collect_selections()` stacks `.selected` over the folds that completed,
with the fold label beside each row, so what every fold chose is one table.
`collect_inner_metrics()` does the same for `.inner_metrics`, so everything
every fold scored is one table. `collect_notes()`, further down, does the
same for `.notes` over every fold.

```{r collect-stacked}
collect_selections(res)

collect_inner_metrics(res)
```

`collect_predictions()` stacks `.predictions` the same way, one row per
held-out row of every completed fold. That table is where a plot of
predicted against observed, or a per-observation loss, would start. It
refuses a run that did not save predictions, naming the control slot to
set. `collect_extracts()` does the same for `.extracts`, one row per fold.

```{r collect-predictions}
collect_predictions(res)
```

`agreement()` counts the selections. Each row is one distinct combination
the folds chose. `n` is how many completed folds chose it, and `prop` is
that count as a share of the completed folds. The most frequent row describes how
stable the tuning procedure's choice was on this data. It is not the final
model's parameters, which come from `nested_final_fit()` running the
procedure once more on the whole dataset.

```{r agreement}
agreement(res)
```

## The plots

`autoplot()` draws the same two facts. The default view is the selections,
one panel per tuned parameter and one point per outer fold.

```{r autoplot-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter, mtry and min_n, with one point per outer fold at the value that fold's inner tuning selected."}
autoplot(res)
```

The other view is the outer scores, one panel per metric and one point per
fold, with a dashed line at the mean that `collect_metrics()` reports.

```{r autoplot-performance, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per metric, rmse and rsq, with one point per outer fold's score on its assessment rows and a dashed line at the mean across folds."}
autoplot(res, type = "performance")
```

## A failed fold

A fold that fails does not end the run. The other folds keep their results,
and the fold that failed is recorded rather than dropped. To show that, the
workflow below adds a range check on horsepower, which refuses to predict
for a car whose horsepower lies outside the range the model was trained on.
One car, the Maserati Bora, has far more horsepower than any other in
`mtcars`. The fold that holds it out cannot score it, so that fold fails.
The chunk mutes tune's progress messages and keeps its warnings.

```{r failed-recipe}
rec <- recipe(mpg ~ ., data = mtcars) |>
  check_range(hp)

wf_checked <- workflow(rec, rf)
```

```{r failed-run, message = FALSE}
set.seed(2)

failed <- nested_tune_grid(wf_checked, folds, grid = grid)

failed
```

The print counts the failure. `.completed` is `FALSE` for the fold that
failed, and its `.notes` say where. `collect_notes()` stacks every fold's
`.notes` into one table with the fold label beside them. The first row for
the failed fold is this package's own note, its `location` naming the
stage. The rows after it are tune's notes about the cause.

```{r failed-notes}
failed_id <- failed |>
  filter(!.completed) |>
  pull(id)

collect_notes(failed) |>
  filter(id == failed_id) |>
  select(location, type, note)
```

A fold can also complete and still carry notes. That happens when an inner
resample failed but tuning still returned a candidate from the resamples
that ran.

On a run with a failed fold, the readers answer over the folds that
completed. `summary()`, `collect_metrics()`, `agreement()`,
`collect_selections()` and `collect_inner_metrics()` each warn once, with
class `nestedtune_partial_summary`, saying how many folds the answer covers.
`collect_notes()` reads every fold and warns about none. `autoplot()` leaves
the failed fold's place on its axis empty.

A run in which no fold completed is a different case. It has no estimate to
report, so every reader that needs a completed fold refuses it:
`collect_metrics()`, `collect_predictions()`, `collect_extracts()`,
`agreement()`, `collect_selections()`, `collect_inner_metrics()`,
`autoplot()` and `nested_final_fit()`. `summary()` and `collect_notes()`
still answer on such a run.

## Subsetting with dplyr

The object is a tibble, so dplyr's verbs work on it. One rule decides what
they hand back. A verb that only reorders the rows, or adds or reorders
columns, returns a `nested_results`. A verb that adds or removes a row, or
drops or overwrites one of the columns above, returns a bare tibble. An
object that no longer holds what the run produced cannot answer for the run.
So adding a column keeps the class.

```{r dplyr-mutate}
with_position <- mutate(res, position = row_number())

class(with_position)
```

Dropping a fold sheds it. The natural case is keeping only the folds that
completed on the run above, which removes one row.

```{r dplyr-filter}
completed_only <- filter(failed, .completed)

class(completed_only)
```

Dropping the columns the run wrote sheds it too, here through base `[`
rather than a dplyr verb, since the same rule governs it.

```{r subset}
class(res[, "id"])
```

Both of those hand back the data and nothing more: a plain tibble print, no
summary of the run, no `collect_metrics()`. A table that has lost a fold, or
lost the columns the run wrote, cannot describe itself as a five-fold
design, so it stops describing itself. To read a partial run, read the run
itself: `collect_metrics(failed)` already averages the folds that completed,
and warns that it did.

Two arguments went unused on this page. `event_level` names which level of
a two-class outcome is the event. `eval_time` names the times at which a
censored-regression run is scored. Both are described on
`?nested_tune_grid`. The second adds an `.eval_time` column to `.metrics`
and to everything `collect_metrics()` reports.
