---
title: "Choosing the inner tuner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the inner tuner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You want the inner search to run some other way than over a fixed grid, or
to score a workflow with nothing to tune. The getting-started guide,
`vignette("nested-cv")`, tunes each outer fold over a fixed grid with
`nested_tune_grid()`. That is one of five ways the inner search can run. The other four are tune's Bayesian optimization,
finetune's two racing methods, and finetune's simulated annealing. Each has
a function in this package that takes the same design and workflow and
returns the same kind of object. This page runs all four on the guide's example and
shows what each fold records about its search. It then scores a workflow
with nothing to tune on the same folds. Last, it runs a baseline and two
tuned workflows through one design in one call with `nested_workflow_map()`.

```{r deps, include = FALSE}
needed <- c("tidymodels", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages and fits random forests",
  "with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

```{r deps-optional, include = FALSE}
has_finetune <- requireNamespace("finetune", quietly = TRUE)
has_race <- has_finetune &&
  all(vapply(
    c("lme4", "BradleyTerry2"),
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
```

## The design and the workflow

These are the guide's, unchanged: five outer folds of `mtcars`, five inner
folds under each, and a random forest with two parameters marked for tuning.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

The grid tuner and the racers score the candidates they are given, the
settings in the grid. The Bayesian and annealing searches propose their own
candidates, and to do that they need each parameter's range. The default
range of `mtry` is not known until the data is seen, so a search over it
fails at inner tuning in every fold. The parameter set below fixes that
range by hand. It also bounds `min_n` to the grid's range, because its
default runs past the size of an inner analysis set here.

```{r param-info}
params <- update(
  extract_parameter_set_dials(wf),
  mtry = mtry(c(2L, 8L)),
  min_n = min_n(c(2L, 10L))
)

params
```

## Bayesian optimization

`nested_tune_bayes()` runs `tune::tune_bayes()` inside each outer fold. The
search starts from a small space-filling set of candidates and scores them
on the inner resamples. It then proposes one new candidate per iteration,
from a Gaussian process fitted to the scores so far. `initial` and `iter`
are the sizes of those two stages, kept small here so the page builds
quickly.

```{r bayes}
set.seed(2)

bayes <- nested_tune_bayes(
  wf,
  folds,
  param_info = params,
  initial = 4,
  iter = 3
)

bayes
```

The print notes that the folds did not search the same grid. Each fold
scored the same number of candidates, but not the same ones, because each
fold proposes candidates from its own scores. With a grid, every fold picks
from the same list, so you can count how many folds chose each candidate.
Here the folds did not all see the same list, and the note says so.

The guide multiplied outer folds by inner resamples by grid rows to count
fits. Here the same count uses `initial + iter` candidates per fold in
place of the grid's rows.

`.inner_metrics` holds what the search inside one fold saw. Here is the
first fold's:

```{r bayes-inner}
bayes$.inner_metrics[[1]]
```

```{r bayes-iter, include = FALSE}
bayes_iter <- range(bayes$.inner_metrics[[1]]$.iter)
```

The `.iter` column says which stage each candidate came from. An `.iter` of
`r bayes_iter[1]` marks the initial set, and the proposals run up to
`r bayes_iter[2]`. A `tune::control_bayes()` passed as `control` reaches the
inner call in every fold. If the control stops a fold's search early, that
fold's `.iter` values stop short of `iter`. `?nested_tune_bayes`
says which of its slots this package sets for itself.

## Racing

The two racing methods in finetune take the same grid the guide uses and
score every candidate on a few inner resamples first. After each further
resample, a candidate that is clearly worse than the current best is
dropped. That saves the fits a full grid search would spend on losing
candidates. `nested_tune_race_anova()` decides with a repeated measures
ANOVA fitted by lme4. `nested_tune_race_win_loss()` decides with a
Bradley-Terry model of pairwise wins fitted by BradleyTerry2. Both refuse at
entry when a package their race needs is not installed.

```{r race-notice, eval = !has_race, echo = FALSE, results = "asis"}
absent <- c("finetune", "lme4", "BradleyTerry2")
absent <- absent[!vapply(
  absent,
  requireNamespace,
  logical(1),
  quietly = TRUE
)]

cat(
  "The racing runs need the **finetune**, **lme4** and **BradleyTerry2**",
  "packages, and",
  if (length(absent) == 1L) "one of them is" else "some of them are",
  "not installed here:",
  paste0(paste0("**", absent, "**", collapse = ", "), "."),
  "The two racing sections",
  if (has_finetune) "have" else "and the simulated annealing section have",
  "been skipped. Install what is missing and rebuild to see them.\n"
)
```

```{r race-anova, eval = has_race}
set.seed(3)

race <- nested_tune_race_anova(wf, folds, grid = grid)

race
```

A race's `.inner_metrics` looks complete but is not. Every candidate in the
grid has a row, yet the rows can rest on different numbers of resamples. The
`n` column says how many inner resamples each candidate was scored on
before it was dropped or the race ended. `collect_inner_metrics()` stacks those tables
with the fold label beside them. So summing `n` over one metric's rows
counts the fits each fold spent on tuning. The fold that spent the fewest is
the one where elimination did the most work:

```{r race-anova-inner, eval = has_race}
fits_per_fold <- collect_inner_metrics(race) |>
  filter(.metric == "rmse") |>
  group_by(id) |>
  summarise(fits = sum(n))

fits_per_fold

cheapest <- slice_min(fits_per_fold, fits, n = 1, with_ties = FALSE)

collect_inner_metrics(race) |>
  filter(id == cheapest$id)
```

```{r race-anova-count, eval = has_race, include = FALSE}
inner_v <- nrow(folds$inner_resamples[[1]])
rmse_rows <- collect_inner_metrics(race) |>
  filter(id == cheapest$id, .metric == "rmse")
eliminated <- sum(rmse_rows$n < inner_v)
```

```{r race-anova-prose, eval = has_race, echo = FALSE, results = "asis"}
cat(sprintf(
  paste0(
    "In `%s`, %d of the %d candidates show `n` below the %d inner ",
    "resamples, so they were dropped before the race ended, and the fold ",
    "spent %d fits on tuning where the full grid costs %d.\n"
  ),
  cheapest$id,
  eliminated,
  nrow(rmse_rows),
  inner_v,
  cheapest$fits,
  inner_v * nrow(grid)
))
```

The win/loss race records the same table, and reads the same way. Its fits
per fold:

```{r race-win-loss, eval = has_race}
set.seed(4)

win_loss <- nested_tune_race_win_loss(wf, folds, grid = grid)

collect_inner_metrics(win_loss) |>
  filter(.metric == "rmse") |>
  group_by(id) |>
  summarise(fits = sum(n))
```

## Simulated annealing

`nested_tune_sim_anneal()` runs `finetune::tune_sim_anneal()` inside each
outer fold. Like the Bayesian search, it proposes its own candidates and
needs the parameter set above. Each proposal is a small random move from
the current candidate, accepted when it scores better and sometimes when it
scores worse. finetune prints a log of every move by default, and the
control passed below keeps the page quiet.

```{r anneal, eval = has_finetune}
set.seed(5)

anneal <- nested_tune_sim_anneal(
  wf,
  folds,
  param_info = params,
  initial = 3,
  iter = 3,
  control = finetune::control_sim_anneal(verbose_iter = FALSE)
)

anneal$.inner_metrics[[1]]
```

```{r anneal-iter, eval = has_finetune, include = FALSE}
anneal_iter <- range(anneal$.inner_metrics[[1]]$.iter)
```

```{r anneal-prose, eval = has_finetune, echo = FALSE, results = "asis"}
cat(sprintf(
  "`.iter` runs from %d, the initial candidates, to %d.\n",
  anneal_iter[1],
  anneal_iter[2]
))
```

Seeding is the same on all five. Seed the session before the call. The same
seed gives the same result serially and in parallel, as
`vignette("nested-cv")` explains under Reproducibility.

## A workflow with nothing to tune

A tuned procedure, the whole resample-tune-select-fit sequence, is usually
compared with something simpler, such as the same model with its parameters
fixed. `nested_fit_resamples()` scores such a workflow on the same nested
design. It runs the same outer loop with the inner stage removed. Each fold
fits the workflow as given on its analysis rows and scores it once on its
assessment rows. `extract_procedure()` on the result says no tuning ran.
What the shared design buys is that the two runs score on identical folds.

```{r baseline}
fixed_rf <- rand_forest(mtry = 2L, min_n = 10L, trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

set.seed(2)
baseline <- nested_fit_resamples(workflow(mpg ~ ., fixed_rf), folds)

extract_procedure(baseline)$tuner
collect_metrics(baseline)
```

Every function that reads a tuned result reads the baseline the same way,
under the same control settings. The one exception is the parameters plot,
which has nothing to draw. The baseline's
`.selected` column holds an empty table on every fold, since nothing was
chosen, so `collect_selections()` and `agreement()` return zero rows. The per-fold metrics join those of a tuned
run by fold label:

```{r baseline-join}
collect_metrics(bayes, summarize = FALSE) |>
  filter(.metric == "rmse") |>
  select(id, tuned = .estimate) |>
  left_join(
    collect_metrics(baseline, summarize = FALSE) |>
      filter(.metric == "rmse") |>
      select(id, fixed = .estimate),
    by = "id"
  )
```

Read the table fold by fold rather than as one difference. Two nested
estimates cannot be subtracted to compare procedures, for the reasons
`vignette("estimate")` gives.

`nested_fit_resamples()` and the five tuning functions do not overlap.
`nested_fit_resamples()` refuses a workflow carrying a `tune()` marker. The
five tuning functions refuse a workflow with none. Each refusal names the
function to use instead.

```{r baseline-refusal, error = TRUE}
nested_fit_resamples(wf, folds)
```

## A set of workflows on one design

A comparison across model families needs every family scored on the same
outer folds. `nested_workflow_map()` takes a `workflow_set()` and the name
of one of the six functions above, `nested_fit_resamples()` included. It
runs each workflow of the set
through that function on one design, so the results come back side by side.
Here the set holds three workflows. They are the random forest above, a
linear model on the same predictors with nothing to tune, and a linear model
on principal components whose count is tuned. The forest's grid goes in the call. The
components workflow cannot use that grid, so its own grid goes in the set's
`option` column with `option_add()`.

```{r wset}
pca <- recipe(mpg ~ ., data = mtcars) |>
  step_pca(all_predictors(), num_comp = tune())

wset <- as_workflow_set(
  forest = wf,
  baseline = workflow(mpg ~ ., linear_reg()),
  components = workflow(pca, linear_reg())
) |>
  option_add(grid = tibble(num_comp = 1:4), id = "components")

set.seed(7)
mapped <- nested_workflow_map(
  wset,
  fn = "nested_tune_grid",
  resamples = folds,
  grid = grid
)

mapped
```

The baseline has nothing to tune. So it ran through `nested_fit_resamples()`
instead of the function that `fn` names, and the print says so beside its
id. Each row's `result` is the object the named function returns for that
workflow. Calling that function by hand, with the same arguments and the
same seed, would give the same object. So each row can be read exactly as
the earlier sections read a single result.

```{r wset-metrics}
collect_metrics(mapped)
```

`collect_metrics()` stacks each workflow's estimate under its id, and the
other collect functions stack their tables the same way. `summary()` summarizes every
workflow, one section per id under one heading for the set.

```{r wset-summary}
summary(mapped)
```

`agreement()` stacks each workflow's selection table under its id, with the
parameter columns ahead of the counts. The baseline tuned nothing and has no
row.

```{r wset-agreement}
agreement(mapped)
```

`autoplot()` has the same two views as on one result. The performance view
puts the workflows along one axis inside a panel per metric. A dashed rule
marks each workflow's `collect_metrics()` mean.

```{r wset-performance, fig.width = 7, fig.height = 3.6, fig.alt = "Two panels, one per metric, with the three workflows along the x axis, one point per outer fold and a dashed rule at each workflow's nested estimate."}
autoplot(mapped, type = "performance")
```

The parameters view keeps the outer folds on the x axis and gives each
workflow's tuned parameter its own panel, labelled by the id. So each
workflow gets its own answer to whether the folds agreed. The baseline has
no panel.

```{r wset-parameters, fig.width = 7, fig.height = 3.6, fig.alt = "One panel per tuned parameter and workflow, labelled by the workflow id, with the outer folds along the x axis and one point per fold at the value it selected."}
autoplot(mapped)
```

What the set does not offer is a ranking of its workflows or a fit of the
best one. Choosing among them by these estimates would be a selection the
outer loop did not nest, as `vignette("estimate")` says. The final fit
for one workflow of the set is `nested_final_fit(mapped, id = "forest")`.
