---
title: "Running the outer loop in parallel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the outer loop in parallel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Nested cross-validation fits many models, and the outer folds are
independent of one another. `nested_tune_grid()` and its siblings run those
folds on [mirai](https://mirai.r-lib.org/) daemons when a pool is connected,
and serially otherwise. Nothing in the call changes between the two. This
page starts a pool of two daemons, runs the getting-started guide's loop on
it, shows the result identical to the same run made serially, and says when
a pool pays. What the call checks before it dispatches, what it sends to
each daemon, and what happens on an interrupt are under the parallel section
of `?nested_tune_grid`.

```{r deps, include = FALSE}
needed <- c("tidymodels", "mirai", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages, runs the outer loop on",
  "**mirai** daemons and fits random forests with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design and the workflow

These are the guide's, unchanged: five outer folds of `mtcars`, five inner
folds under each, a random forest with two parameters marked for tuning, and
a grid of six candidates. `vignette("nested-cv")` walks through each of them.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

## The pool

Parallelism is switched on by connecting daemons and off by disconnecting
them. There is no argument on the loop functions for it. `mirai::daemons()`
starts the pool, and `mirai::status()` reports how many daemons are
connected. Two or more is the threshold, the same one tune uses.

```{r daemons}
mirai::daemons(2)

mirai::status()$connections
```

Each daemon is a separate R process. It loads nestedtune and the packages
the workflow needs from an installed library, and the first parallel call
after starting a pool is the slow one, since it is what makes each daemon
load the tidymodels stack.

## The same call

With a pool connected the outer folds are dispatched to the daemons, and
each fold's inner tuning runs serially on its daemon, because parallelism
inside a fold on top of parallelism across folds would oversubscribe the
cores.

```{r run-parallel}
set.seed(2)

par_res <- nested_tune_grid(wf, folds, grid = grid)

par_res
```

The result carries no mark of how it was computed. `mirai::daemons(0)`
disconnects the daemons, and after it the same call runs serially.

```{r daemons-stop}
mirai::daemons(0)

mirai::status()$connections
```

## The same result

Each outer fold's seeds are drawn from the session state at entry, fixed by
the fold's position in the design, before anything is dispatched, as
`vignette("nested-cv")` explains under Reproducibility. So the serial run
under the same seed is the same procedure on the same resamples.

```{r run-serial}
set.seed(2)

ser_res <- nested_tune_grid(wf, folds, grid = grid)
```

```{r identity}
same <- c(
  metrics = identical(par_res$.metrics, ser_res$.metrics),
  selected = identical(par_res$.selected, ser_res$.selected),
  tuning_seed = identical(par_res$.tuning_seed, ser_res$.tuning_seed)
)

same
```

`identical()` on the outer scores, the selections and the tuning seeds
returns `r all(same)` for each, and which run used daemons is not
recoverable from either. That is what lets a pool be a matter of the session
rather than of the analysis: a script gives the same answer on a laptop with
no daemons and on a workstation with many.

## A pool's payoff

`mtcars` has `r nrow(mtcars)` rows and the grid has `r nrow(grid)` points,
so each fold takes a fraction of a second, and two daemons loading the
tidymodels stack cost more than the folds save. This page shows the parallel
path working and its result matching the serial one, not a speedup. The
speedup arrives where the folds are expensive: larger data, a bigger grid, a
slower engine, or preprocessing the loop has to redo in every fold. The fit
count in `vignette("nested-cv")` says where the cost lives, and what is
parallel here is the outer loop over that cost.
