---
title: "Running the outer loop in parallel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the outer loop in parallel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A nested run fits many models, and you can make it finish sooner. The
outer folds do not depend on one another, so they can run at the same time.
`nested_tune_grid()` and its siblings run them on
[mirai](https://mirai.r-lib.org/) daemons, separate R processes, when a
pool of daemons is connected. Without a pool, the folds run one after
another. Nothing in the call changes between the two.

This page starts a pool of two daemons and runs the getting-started guide's
loop on it. It then shows that the result is identical to the same run made
serially. Last, it says when a pool pays. The parallel section of
`?nested_tune_grid` says what the call checks before it dispatches, what it
sends to each daemon, and what happens on an interrupt.

```{r deps, include = FALSE}
needed <- c("tidymodels", "mirai", "ranger")
missing_pkgs <- needed[
  !vapply(needed, requireNamespace, logical(1), quietly = TRUE)
]
has_deps <- length(missing_pkgs) == 0L
```

```{r deps-notice, eval = !has_deps, echo = FALSE, results = "asis"}
cat(
  "This page attaches the **tidymodels** packages, runs the outer loop on",
  "**mirai** daemons and fits random forests with the **ranger** package.",
  if (length(missing_pkgs) == 1L) {
    sprintf("The **%s** package is not installed here,", missing_pkgs)
  } else {
    sprintf(
      "The **%s** packages are not installed here,",
      paste(missing_pkgs, collapse = "** and **")
    )
  },
  "so the rest of this page has been skipped. Install",
  if (length(missing_pkgs) == 1L) "it" else "them",
  "and rebuild to see it.\n"
)

knitr::knit_exit()
```

```{r libraries, message = FALSE}
library(tidymodels)
library(nestedtune)
```

## The design and the workflow

These are the guide's, unchanged. There are five outer folds of `mtcars`,
five inner folds under each, and a random forest with two parameters marked
for tuning. The grid holds six candidates, the settings to try.
`vignette("nested-cv")` walks through each of them.

```{r design}
set.seed(1)

folds <- nested_resamples(
  mtcars,
  outside = vfold_cv(v = 5),
  inside = vfold_cv(v = 5)
)

rf <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

wf <- workflow(mpg ~ ., rf)

grid <- expand.grid(mtry = c(2L, 5L, 8L), min_n = c(2L, 10L))
```

## The pool

You switch parallelism on by connecting daemons and off by disconnecting
them. There is no argument on the loop functions for it. `mirai::daemons()`
starts the pool, and `mirai::status()` reports how many daemons are
connected. With two or more connected, the folds go out to the daemons.
With fewer, they run in the session, the same rule tune uses.

```{r daemons}
mirai::daemons(2)

mirai::status()$connections
```

Each daemon is a separate R process. It loads nestedtune and the packages
the workflow needs from an installed library. So the first parallel call
after starting a pool is the slow one, because it makes each daemon load the
tidymodels stack.

## The same call

With a pool connected, the outer folds are sent to the daemons. Each fold's
inner tuning runs serially on its daemon. Parallelism inside a fold on top
of parallelism across folds oversubscribes the cores.

```{r run-parallel}
set.seed(2)

par_res <- nested_tune_grid(wf, folds, grid = grid)

par_res
```

The result carries no mark of how it was computed. `mirai::daemons(0)`
disconnects the daemons, and after it the same call runs serially.

```{r daemons-stop}
mirai::daemons(0)

mirai::status()$connections
```

## The same result

Each outer fold's seeds are drawn from the session state at entry, before
anything is dispatched. A fold's seeds depend on its position in the
design, as `vignette("nested-cv")` explains under Reproducibility. So a
serial run under the same seed repeats the parallel run: the same
procedure, the same tune-and-fit steps, on the same resamples.

```{r run-serial}
set.seed(2)

ser_res <- nested_tune_grid(wf, folds, grid = grid)
```

```{r identity}
same <- c(
  metrics = identical(par_res$.metrics, ser_res$.metrics),
  selected = identical(par_res$.selected, ser_res$.selected),
  tuning_seed = identical(par_res$.tuning_seed, ser_res$.tuning_seed)
)

same
```

`identical()` on the outer scores, the selections and the tuning seeds
returns `r all(same)` for each. Neither result records whether daemons
were used, so a script gives the same answer on a laptop with no daemons
and on a workstation with many.

## A pool's payoff

`mtcars` has `r nrow(mtcars)` rows and the grid has `r nrow(grid)` points,
so each fold takes a fraction of a second. Two daemons loading the
tidymodels stack cost more than the folds save. This page shows the parallel
path working and its result matching the serial one, not a speedup. The
speedup arrives where the folds are expensive: larger data, a bigger grid, a
slower engine, or preprocessing the loop has to redo in every fold. The fit
count in `vignette("nested-cv")` says where the cost lives. A pool splits
that cost across the outer folds.
